Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-11-01 Manuscript/Manuscript.Rmd
Add popularity data and missing data pattern plot

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
library(tidyverse)
library(mice)
library(metamisc)
```

@@

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(pat = rep("Systematic", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
  tibble(pat = rep("Sporadic", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
  mutate(across(everything(), as.factor))
ggplot(patterns, aes(x = name, y = rownr, fill = value, type = pat)) +
  geom_tile(color = "black") +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~pat) +
  theme_void() +
  theme(legend.position = "none") 
```

-	ADD: missingness mech here

@@

@@

- We'll use the IMPACT data (`metamisc::impact`) and simulated MNAR data (based on `mice::popularity`).

- IMPACT: A data frame with 11022 observations on the following 11 variables:
`name` Name of the study,
`type` Type of study, RCT: randomized controlled trial,OBS: observational cohort,
`age` Age of the patient,
`motor_score` Glasgow Coma Scale motor score,
`pupil` Pupillary reactivity,
`ct` Marshall Computerized Tomography classification,
`hypox` Hypoxia (0=no, 1=yes),
`hypots` Hypotension (0=no, 1=yes),
`tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
`edh` Epidural hematoma (0=no, 1=yes),
`mort` 6-month mortality (0=alive, 1=dead).

```{r impact}
# load data
data("impact")
# descriptive statistics
psych::describe(impact)[,c(2:5,8:9)]
```

>>>>

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')

# packages
library(tidyverse)
library(mice)
library(miceadds)
library(metamisc)
library(pan)

# functions
miceadds::source.all(path = "../R")

# plot parameters
plot_col <- mice:::mdc(1:2) %>% setNames(c("observed", "missing"))
```

@@

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(
    pat = rep("systematic", 5),
    rownr = 5:1,
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(X1, X2, X3)),
  tibble(
    pat = rep("sporadic", 5),
    rownr = 5:1,
    X1 = c("observed", "missing", rep("observed", 3)),
    X2 = c(rep("observed", 2), rep("missing", 3)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(X1, X2, X3))
) %>%
  mutate(across(everything(), as.factor),
         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE))
# plot
ggplot(patterns, aes(
  x = name,
  y = rownr,
  fill = value,
  type = pat
)) +
  geom_tile(color = "black") +
  scale_y_discrete(labels = c(2, 2, 1, 1, 1)) +
  scale_fill_manual(values = plot_col, name = "Datapoint") +
  facet_wrap( ~ pat, ncol = 2) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "Variable", y = "Cluster")
```

-	What kinds of missingness are there? ADD: missingness mechanisms here.

@@

- Additional difficulty that is addressed in this tutorial: MNAR data.

@@

- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)).

- `impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:
    * `name` Name of the study,
    * `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
    * `age` Age of the patient,
    * `motor_score` Glasgow Coma Scale motor score,
    * `pupil` Pupillary reactivity,
    * `ct` Marshall Computerized Tomography classification,
    * `hypox` Hypoxia (0=no, 1=yes),
    * `hypots` Hypotension (0=no, 1=yes),
    * `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
    * `edh` Epidural hematoma (0=no, 1=yes),
    * `mort` 6-month mortality (0=alive, 1=dead).

```{r impact}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# missingness
md_pat(impact)
```
-> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).

- `popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables: 
    * `pupil`	Pupil number within class,
    * `class`	Class number,
    * `extrav`	Pupil extraversion,
    * `sex`	Pupil gender,
    * `texp`	Teacher experience (years),
    * `popular`	Pupil popularity,
    * `popteach`	Teacher popularity.

```{r pop}
# load data
pop <- readRDS("../Data/popNCR.RDS")
# missingness
md_pat(pop)
```

### 2021-11-01 Manuscript/Manuscript.tex
Clean up repository, updat readme
Add popularity data and missing data pattern plot

\begin{center}\includegraphics{Multilevel-Missing-Data_files/figure-latex/patterns-1} \end{center}

@@

\begin{itemize}
\item
  ADD: missingness mech here
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What types of multilevel methods are available? General overview of
  approaches, see \citet{audi18} \citet{grun18}. E.g., imputation of
  study level versus patient-level covariates, and one-stage imputation
  versus two-stage imputation methods.
\end{itemize}

@@

\begin{itemize}
\item
  We'll use the IMPACT data (\texttt{metamisc::impact}) and simulated
  MNAR data (based on \texttt{mice::popularity}).
\item
  IMPACT: A data frame with 11022 observations on the following 11
  variables: \texttt{name} Name of the study, \texttt{type} Type of
  study, RCT: randomized controlled trial,OBS: observational cohort,
  \texttt{age} Age of the patient, \texttt{motor\_score} Glasgow Coma
  Scale motor score, \texttt{pupil} Pupillary reactivity, \texttt{ct}
  Marshall Computerized Tomography classification, \texttt{hypox}
  Hypoxia (0=no, 1=yes), \texttt{hypots} Hypotension (0=no, 1=yes),
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  \texttt{edh} Epidural hematoma (0=no, 1=yes), \texttt{mort} 6-month
  mortality (0=alive, 1=dead).
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # load data
R> data("impact")
R> # descriptive statistics
R> psych::describe(impact)[,c(2:5,8:9)]
\end{CodeInput}
\begin{CodeOutput}
                 n  mean    sd median min max
name*        11022  8.32  4.37      8   1  15
type*        11022  1.73  0.44      2   1   2
age          11022 34.93 15.89     31  14  93
motor_score* 11022  2.59  1.18      3   1   4
pupil*       11022  1.46  0.71      1   1   3
ct*          11022  1.98  0.89      2   1   3
hypox        11022  0.22  0.41      0   0   1
hypots       11022  0.17  0.38      0   0   1
tsah         11022  0.45  0.50      0   0   1
edh          11022  0.13  0.34      0   0   1
mort         11022  0.26  0.44      0   0   1
\end{CodeOutput}
\end{CodeChunk}

>>>>

\begin{center}\includegraphics{Manuscript_files/figure-latex/patterns-1} \end{center}

@@

\begin{itemize}
\item
  What kinds of missingness are there? ADD: missingness mechanisms here.
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What types of multilevel methods are available? General overview of
  approaches, see \citet{audi18} \citet{grun18}. E.g., imputation of
  study level versus patient-level covariates, and one-stage imputation
  versus two-stage imputation methods.
\item
  Additional difficulty that is addressed in this tutorial: MNAR data.
\end{itemize}

@@

\begin{itemize}
\item
  We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
  version of the \texttt{mice::popmis} data (i.e., a variation on the
  Hox (2010) popularity data, where the missingness in the variables is
  either missing at random (MAR) or missing not at random (MNAR)).
\item
  \texttt{impact} is traumatic brain injury data with patients clustered
  in studies, \(n_{\text{participants}} = 11022\) and
  \(n_{\text{clusters}} = 15\), on the following 11 variables:

  \begin{itemize}
  \tightlist
  \item
    \texttt{name} Name of the study,
  \item
    \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
    observational cohort),
  \item
    \texttt{age} Age of the patient,
  \item
    \texttt{motor\_score} Glasgow Coma Scale motor score,
  \item
    \texttt{pupil} Pupillary reactivity,
  \item
    \texttt{ct} Marshall Computerized Tomography classification,
  \item
    \texttt{hypox} Hypoxia (0=no, 1=yes),
  \item
    \texttt{hypots} Hypotension (0=no, 1=yes),
  \item
    \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  \item
    \texttt{edh} Epidural hematoma (0=no, 1=yes),
  \item
    \texttt{mort} 6-month mortality (0=alive, 1=dead).
  \end{itemize}
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # load data
R> data("impact")
R> # # descriptive statistics
R> # by(impact, impact$name, summary) 
R> # psych::describe(impact)[,c(2:5,8:9)]
R> # missingness
R> md_pat(impact)
\end{CodeInput}
\begin{CodeOutput}
 /\     /\
{  `---'  }
{  O   O  }
==>  V <==  No need for mice. This data set is completely observed.
 \  \|/  /
  `-----'
\end{CodeOutput}
\begin{CodeOutput}
      name type age motor_score pupil ct hypox hypots tsah edh mort  
11022    1    1   1           1     1  1     1      1    1   1    1 0
         0    0   0           0     0  0     0      0    0   0    0 0
\end{CodeOutput}
\end{CodeChunk}

-\textgreater{} Why are there no missings? According to the
\href{https://cran.r-project.org/web/packages/metamisc/metamisc.pdf}{vignette},
the data is already imputed (Steyerberg et al, 2008).

\begin{itemize}
\tightlist
\item
  \texttt{popNCR} is a simulated dataset with pupils clustered in
  classes, \(n_{\text{participants}} = 2000\),
  \(n_{\text{clusters}} = 100\), on 7 variables:

  \begin{itemize}
  \tightlist
  \item
    \texttt{pupil} Pupil number within class,
  \item
    \texttt{class} Class number,
  \item
    \texttt{extrav} Pupil extraversion,
  \item
    \texttt{sex} Pupil gender,
  \item
    \texttt{texp} Teacher experience (years),
  \item
    \texttt{popular} Pupil popularity,
  \item
    \texttt{popteach} Teacher popularity.
  \end{itemize}
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # load data
R> pop <- readRDS("../Data/popNCR.RDS")
R> # missingness
R> md_pat(pop)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-1} \end{center}
